PANNEURO_analysis

### 2020-11-05 PANNEURO_analysis_WHO_pres.Rmd
alish sensitivity analyses

---
title: "PANNEURO_analysis"
author: "Christopher Huebel, Alish Palmos, Molly Davies, and Laura Hegemann"
date: "Sys.Date()"
output:
  html_document:
    df_print: paged
    highlight: monochrome
    number_sections: no
    theme: cerulean
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
    toc_depth: '2'
  word_document:
    toc: yes
    toc_depth: '2'
html_notebook:
  theme: cerulean
code_folding: show
toc: yes
---

@@

#Packages
Install packages (if they are not available in your version of R)
```{r Installing packages, include = FALSE}
#install.packages("summarytools")
#install.packages("tidyverse")
#install.packages("psych")
#install.packages("broom")
#install.packages("skimr")
#install.packages("polycor")
#install.packages("corrplot")
#install.packages("patchwork")
#instlal.packages("ggpubr")
```

Load packages
```{r Load packages, include = FALSE}
library(knitr)
library(summarytools)
library(psych)
library(polycor)
library(corrplot)
library(patchwork)
library(broom)
library(tidyverse)
library(ggpubr)
```

# Colour palettes
Define colours for plotting this are the standard coping colours
```{r Colour palettes: COPING, include = FALSE}
COPINGpalette2 <- c("#78D9C5",
                    "#F5BE5E")

@@

# ggplot theme
Set up ggplot theme for the plots
```{r ggplot theme, include=FALSE}
theme_personal <-  theme(
    text = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    axis.text = element_text(color = "black"),
    axis.title.y = element_blank(), 
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    panel.background = element_blank(), 
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(
      colour = "gray",
      linetype = "dashed",
      size = 0.2
      ),
    axis.ticks = element_blank()
    )
```

@@

```{r source file path}
#source data directory: data_path
source("../PANCHANGE_raw_path.R")
```

## Read in data
```{r read in data, include = FALSE}
dat.raw <- readRDS(file = paste0(data_path, "/baseline_and_cfs_wave6.rds")) #NEEDS CHANGING
dim(dat.raw)
colnames(dat.raw)
```

@@

```{r bmi recode}

@@

```{r key worker status frequencies by probable Covid}
key_worker_category_per_Covid_prop_total <- dat.raw %>%
  filter(!is.na(key_worker)) %>%
  filter(!is.na(covid_probable_case_incl_screening)) %>%
  count(key_worker, covid_probable_case_incl_screening) %>%
  mutate(Prop = round(n/sum(n), 2))

@@

# Scoring for neuro measures

# ALSfrs (not included)
```{r ALSfrs col names, include = FALSE}
als.col.names <- dat.raw %>%
  select(
    starts_with("alsfrs") & ends_with("numeric")
  ) %>%
  colnames()

@@

```{r ALSfrs row sums, include = FALSE}
dat.raw <- dat.raw %>%
  mutate(als.sum.score =
  rowSums(select(.,als.col.names), na.rm = F)
   )
```

```{r ALSfrs frequencies, include = FALSE}
dat.raw %>%
  filter(!is.na(startDate.coping)) %>%
  filter(!is.na(covid_probable_case_incl_screening)) %>%
  filter(!is.na(als.sum.score)) %>%
  group_by(covid_probable_case_incl_screening) %>%
  count(covid_probable_case_incl_screening, als.sum.score) %>%
  mutate(Prop = round(n/sum(n), 2))

@@

# ESS
```{r ESS col names, include = FALSE}
ess.col.names <- dat.raw %>%
  select(
    starts_with("ess") & ends_with("numeric")
  ) %>%
  colnames()

ess.col.names
```

```{r ESS row sums, include = FALSE}

# Create vector with scoring key: If the item is reversed coded use -1. 
# There needs to be the correct number of values for each item in the questionnaire. 
ess.items_key <- c(1,1,1,1,1,1,1,1)

# Create vector with items that are included in sum score
ess.items <- c(
  "ess.sitting_and_reading_numeric",
  "ess.watching_tv_numeric",
  "ess.sitting_inactive_in_a_public_place_numeric",
  "ess.break_passenger_car_hour_numeric",
  "ess.afternoon_rest_circumstances_lying_numeric",
  "ess.sitting_and_talking_to_someone_numeric",
  "ess.sitting_quietly_after_a_lunch_without_alcohol_numeric",
  "ess.car_minutes_stopped_traffic_numeric"
  )

# Calculate sum score
ess.scored_items <- scoreItems(keys = ess.items_key,
                                        items = dat.raw[ess.items],
                                        totals = TRUE, 
                                        missing = TRUE, 
                                        impute = 'none', 
                                        min = 0, 
                                        max = 3)
# Add in column to tibble
dat.raw$ess.sum_score_raw <- ess.scored_items$scores

# Look at values
dat.raw %>%
  group_by(covid_probable_case_incl_screening) %>% 
  descr(ess.sum_score_raw,
        stats = "common"
        )

@@

```{r ESS raw scores by Covid case}
ess_frequencies <- dat.raw %>%
  filter(!is.na(startDate.coping)) %>%
  filter(!is.na(covid_probable_case_incl_screening)) %>%
  filter(!is.na(ess.sum_score_raw)) %>%
  group_by(covid_probable_case_incl_screening) %>%
  count(covid_probable_case_incl_screening, ess.sum_score_raw) %>%
  mutate(Prop = round(n/sum(n), 2))

kable(ess_frequencies)
```

ESS scores: Interpretation
•	0 to 10 = normal range of sleepiness in healthy adults
•	11 to 14 = mild sleepiness
•	15 to 17 = moderate sleepiness
•	18 to 24 = severe sleepiness

```{r ess clinical categories}
dat.raw <- dat.raw %>% 
  mutate(ess.cutoff.scores =
           case_when(ess.sum_score_raw <= 10 ~ "Normal sleepiness",
                     ess.sum_score_raw > 10 & ess.sum_score_raw <= 14 ~ "Mild sleepiness",
                     ess.sum_score_raw > 14 & ess.sum_score_raw <= 17 ~ "Moderate sleepiness",
                     ess.sum_score_raw > 17 & ess.sum_score_raw <= 24 ~ "Severe sleepiness",
                     )
           )

```{r ESS clinical categories frequencies}
ess_frequencies_clinical <- dat.raw %>%
  filter(!is.na(startDate.coping)) %>%
  filter(!is.na(covid_probable_case_incl_screening)) %>%
  filter(!is.na(ess.sum_score_raw)) %>%
  group_by(covid_probable_case_incl_screening) %>%
  count(covid_probable_case_incl_screening, ess.cutoff.scores) %>%
  mutate(Prop = round(n/sum(n), 2))

kable(ess_frequencies_clinical)
```

# NMSW
```{r nmsw col names, include = FALSE}
nmsw.col.names <- dat.raw %>%
  select(
    starts_with("nmsw") & ends_with("numeric")
  ) %>% 
  colnames()
nmsw.col.names

```{r scoring NMSW, include = FALSE}
dat.raw <- dat.raw %>% 
  mutate(nmsw.sum_score = 
           rowSums(select(.,nmsw.col.names), na.rm = FALSE)
           )

@@

```{r select cfs variables, include = FALSE}
cfs.col.names <- dat.raw %>%
  select(
    starts_with("cfs") & ends_with("numeric")
  ) %>%
  colnames()
cfs.col.names
```

```{r calculate sum score, include = FALSE}
dat.raw <- dat.raw %>%
  mutate(cfs.sum_score =
   rowSums(select(.,cfs.col.names), na.rm = F) - 11
   )
cfs.col.names.physio <- cfs.col.names[1:7]

@@

```{r get sum scores by probable Covid}
dat.raw %>%
  filter(!is.na(startDate.coping)) %>%
  filter(!is.na(covid_probable_case_incl_screening)) %>%
  filter(!is.na(cfs.sum_score)) %>%
  group_by(covid_probable_case_incl_screening) %>%
  summarise(mean_total = mean(cfs.sum_score, na.rm = TRUE),
            mean_physical = mean(cfs.sum_score_physical, na.rm = TRUE),
            mean_psychological = mean (cfs.sum_score_psyhchological, na.rm = TRUE))

@@

CFS cut-off scores require individual symptom-level scores to be binary.
Script below recodes variables to binary.
```{r recode cfs variables to binary, include = FALSE}
dat.raw$cfs.correct_word_find_difficult_binary <- if_else(dat.raw$cfs.correct_word_find_difficult_numeric >= 3, 1,0)
dat.raw$cfs.do_you_have_problems_with_tiredness_binary<- if_else(dat.raw$cfs.do_you_have_problems_with_tiredness_numeric >= 3, 1,0)
dat.raw$cfs.do_you_feel_sleepy_or_drowsy_binary <- if_else(dat.raw$cfs.do_you_feel_sleepy_or_drowsy_numeric >= 3, 1,0)
dat.raw$cfs.do_you_feel_weak_binary <- if_else(dat.raw$cfs.do_you_feel_weak_numeric >= 3, 1,0)
dat.raw$cfs.do_you_have_difficulty_concentrating_binary <- if_else(dat.raw$cfs.do_you_have_difficulty_concentrating_numeric >= 3, 1,0)
dat.raw$cfs.do_you_have_problems_starting_things_binary <- if_else(dat.raw$cfs.do_you_have_problems_starting_things_numeric >= 3, 1,0)
dat.raw$cfs.do_you_lack_energy_binary <- if_else(dat.raw$cfs.do_you_lack_energy_numeric >= 3, 1,0)
dat.raw$cfs.do_you_need_to_rest_more_binary <- if_else(dat.raw$cfs.do_you_need_to_rest_more_numeric >= 3, 1,0)
dat.raw$cfs.how_is_your_memory_numeric_binary <- if_else(dat.raw$cfs.how_is_your_memory_numeric >= 3, 1,0) 
dat.raw$cfs.make_slips_speaking_tongue_binary <- if_else(dat.raw$cfs.make_slips_speaking_tongue_numeric >= 3, 1,0)
dat.raw$cfs.strength_muscles_binary <- if_else(dat.raw$cfs.strength_muscles_numeric >= 3, 1,0)
```

```{r cfs select binary variables, include = FALSE}
cfs.col.names.bi <- dat.raw %>%
  select(
    starts_with("cfs") & ends_with("binary")
  ) %>%
  colnames()
cfs.col.names.bi
```

```{r calculate cfs cut-off groups, include = FALSE}
dat.raw <- dat.raw %>%
  mutate(cfs.sum_score_binary =
   rowSums(select(.,cfs.col.names.bi), na.rm = F) 
   )

@@

```{r cfs fatigue cases by probable covid}
dat.raw %>%
  filter(!is.na(startDate.coping)) %>%
  filter(!is.na(covid_probable_case_incl_screening)) %>%
  filter(!is.na(cfs.sum_score_case)) %>%
  group_by(covid_probable_case_incl_screening) %>%
  count(covid_probable_case_incl_screening, cfs.sum_score_case) %>%
  mutate(Prop = round(n/sum(n), 2))
```

```{r cfs fatigue cases by probable covid by Sample}
fatigue_by_covid_by_sample <- dat.raw %>%
  filter(!is.na(startDate.coping)) %>%
  filter(!is.na(covid_probable_case_incl_screening)) %>%
  filter(!is.na(cfs.sum_score_case)) %>%
  group_by(covid_probable_case_incl_screening, Sample) %>%
  count(covid_probable_case_incl_screening, cfs.sum_score_case) %>%
  mutate(Prop = round(n/sum(n), 2))

kable(fatigue_by_covid_by_sample)
```

## Memory

>>>>

---
title: "PANNEURO_analysis"
author: "Christopher Huebel"
date: "Sys.Date()"
output:
  html_document:
    df_print: paged
    highlight: monochrome
    number_sections: no
    theme: cerulean
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
  word_document:
    toc: yes
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: '2'
html_notebook:
  theme: cerulean
code_folding: show
toc: yes
---

@@

#Packages
Install packages (if they are not available in your version of R)
```{r Installing packages}
#install.packages("summarytools")
#install.packages("tidyverse")
#install.packages("psych")
#install.packages("broom")
#install.packages("skimr")
#install.packages("polycor")
#install.packages("corrplot")
#install.packages("patchwork")
#instlal.packages("ggpubr")
```

Load packages
```{r Load packages}
library(knitr)
library(summarytools)
library(psych)
library(polycor)
library(corrplot)
library(patchwork)
library(broom)
library(tidyverse)
library(ggpubr)
```

# Colour palettes
Define colours for plotting this are the standard coping colours
```{r Colour palettes: COPING}
COPINGpalette2 <- c("#78D9C5",
                    "#F5BE5E")

@@

# ggplot theme
Set up ggplot theme for the plots
```{r ggplot theme}
theme_personal <-  theme(
    text = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    axis.text = element_text(color = "black"),
    axis.title.y = element_blank(), 
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    panel.background = element_blank(), 
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(
      colour = "gray",
      linetype = "dashed",
      size = 0.2
      ),
    axis.ticks = element_blank()
    )
```

@@

```{r source file path}
#source data directory: data_path
source("/Users/alishpalmos/Documents/GitHub//PANCHANGE_raw_path.R")
```

## Read in data
```{r read in data}
dat.raw <- readRDS(file = paste0(data_path, "/baseline_and_cfs_wave6.rds")) #NEEDS CHANGING
dim(dat.raw)
colnames(dat.raw)
```

@@

@@

```{r}
key_worker_category_per_Covid_prop_total <- dat.raw %>%
  filter(!is.na(key_worker)) %>%
  filter(!is.na(covid_probable_case_incl_screening)) %>%
  count(key_worker, covid_probable_case_incl_screening) %>%
  mutate(Prop = round(n/sum(n), 2))

@@

**ADD SCORING FOR NEURO MEASURES**
# ALSfrs
```{ALSfrs col names r}
als.col.names <- dat.raw %>%
  select(
    starts_with("alsfrs") & ends_with("numeric")
  ) %>%
  colnames()

@@

```{ALSfrs row sums r}
dat.raw <- dat.raw %>%
  mutate(als.sum.score =
  rowSums(select(.,als.col.names), na.rm = F)
   )
```

```{ALSfrs frequencies r}
dat.raw %>%
  filter(!is.na(startDate.coping)) %>%
  filter(!is.na(covid_probable_case_incl_screening)) %>%
  filter(!is.na(als.sum.score)) %>%
  group_by(covid_probable_case_incl_screening) %>%
  count(covid_probable_case_incl_screening, als.sum.score) %>%
  mutate(Prop = round(n/sum(n), 2))

```
# ESS
```{ESS col names r}
ess.col.names <- dat.raw %>%
  select(
    starts_with("ess") & ends_with("numeric")
  ) %>%
  colnames()

@@

```{ESS row sums r}
dat.raw <- dat.raw %>%
  mutate(ess.col.names =
   rowSums(select(.,ess.col.names), na.rm = F)
   )
```

```{ESS frequencies r}
dat.raw %>%
  filter(!is.na(startDate.coping)) %>%
  filter(!is.na(covid_probable_case_incl_screening)) %>%
  filter(!is.na(ess.col.names)) %>%
  group_by(covid_probable_case_incl_screening) %>%
  count(covid_probable_case_incl_screening, ess.col.names) %>%
  mutate(Prop = round(n/sum(n), 2))
```

```{r nmsw col names}
nmsw.col.names <- dat.raw %>%
  select(
    starts_with("nmsw") & ends_with("numeric")
  ) %>% 
  colnames()
nmsw.col.names

@@

```{r scoring NMSW}
dat.raw <- dat.raw %>% 
  mutate(nmsw.sum_score = 
           rowSums(select(.,nmsw.col.names), na.rm = FALSE)
           )

@@

```{r}
cfs.col.names <- dat.raw %>%
  select(
    starts_with("cfs") & ends_with("numeric")
  ) %>%
  colnames()
cfs.col.names
```
```{r}
dat.raw <- dat.raw %>%
  mutate(cfs.sum_score =
   rowSums(select(.,cfs.col.names), na.rm = F) - 11
   )
cfs.col.names.physio <- cfs.col.names[1:7]

@@

```{r}
dat.raw %>%
  filter(!is.na(startDate.coping)) %>%
  filter(!is.na(covid_probable_case_incl_screening)) %>%
  filter(!is.na(cfs.sum_score)) %>%
  group_by(covid_probable_case_incl_screening) %>%
  summarise(mean_total = mean(cfs.sum_score, na.rm = TRUE),
            mean_physical = mean(cfs.sum_score_physical, na.rm = TRUE),
            mean_psychological = mean (cfs.sum_score_psyhchological, na.rm = TRUE))

@@

```{r}
dat.raw$cfs.correct_word_find_difficult_binary <- if_else(dat.raw$cfs.correct_word_find_difficult_numeric >= 3, 1,0)
dat.raw$cfs.do_you_have_problems_with_tiredness_binary<- if_else(dat.raw$cfs.do_you_have_problems_with_tiredness_numeric >= 3, 1,0)
dat.raw$cfs.do_you_feel_sleepy_or_drowsy_binary <- if_else(dat.raw$cfs.do_you_feel_sleepy_or_drowsy_numeric >= 3, 1,0)
dat.raw$cfs.do_you_feel_weak_binary <- if_else(dat.raw$cfs.do_you_feel_weak_numeric >= 3, 1,0)
dat.raw$cfs.do_you_have_difficulty_concentrating_binary <- if_else(dat.raw$cfs.do_you_have_difficulty_concentrating_numeric >= 3, 1,0)
dat.raw$cfs.do_you_have_problems_starting_things_binary <- if_else(dat.raw$cfs.do_you_have_problems_starting_things_numeric >= 3, 1,0)
dat.raw$cfs.do_you_lack_energy_binary <- if_else(dat.raw$cfs.do_you_lack_energy_numeric >= 3, 1,0)
dat.raw$cfs.do_you_need_to_rest_more_binary <- if_else(dat.raw$cfs.do_you_need_to_rest_more_numeric >= 3, 1,0)
dat.raw$cfs.how_is_your_memory_numeric_binary <- if_else(dat.raw$cfs.how_is_your_memory_numeric >= 3, 1,0) 
dat.raw$cfs.make_slips_speaking_tongue_binary <- if_else(dat.raw$cfs.make_slips_speaking_tongue_numeric >= 3, 1,0)
dat.raw$cfs.strength_muscles_binary <- if_else(dat.raw$cfs.strength_muscles_numeric >= 3, 1,0)
```

```{r}
cfs.col.names.bi <- dat.raw %>%
  select(
    starts_with("cfs") & ends_with("binary")
  ) %>%
  colnames()
cfs.col.names.bi
```
```{r}
dat.raw <- dat.raw %>%
  mutate(cfs.sum_score_binary =
   rowSums(select(.,cfs.col.names.bi), na.rm = F) 
   )

@@

```{r}
dat.raw %>%
  filter(!is.na(startDate.coping)) %>%
  filter(!is.na(covid_probable_case_incl_screening)) %>%
  filter(!is.na(cfs.sum_score_case)) %>%
  group_by(covid_probable_case_incl_screening) %>%
  count(covid_probable_case_incl_screening, cfs.sum_score_case) %>%
  mutate(Prop = round(n/sum(n), 2))
```
